An Innovative Storage Stack Addressing Extreme Scale Platforms and Big Data Applications

### 2014-05-12 paper.tex
near final edit and incorporated the performance graphs

\documentclass[conference]{IEEEtran}
\pdfpagewidth=8.5in
\pdfpageheight=11in
%\usepackage{subfig}
%\usepackage{subfigure}
\usepackage[pdftex]{}
\usepackage{graphicx}
%\usepackage{todonotes}
\usepackage{listings}
\lstset {% general command to set parameter(s)
         language=C,
     basicstyle=\footnotesize,               % print whole listing small
%     keywordstyle=\color{black}\bfseries, % underlined bold black keywords
%     identifierstyle =\color{black},  % nothing happens
%     commentstyle=\color{black}\emph, % white comments
     %stringstyle=\ttfamily,          % typewriter type for strings
%     stringstyle=\color{black},       % typewriter type for strings
         tabsize=4,
         showtabs=false,
     showstringspaces=false}
%can't figure this one out for particles bandwidth
%\usepackage[caption,label]{subfig}

@@

\addtolength{\parskip}{-0.06in}

@@

\begin{abstract}
The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
project is going to have significant impact on storage systems design within
and beyond the HPC community. With phase 1 of the project complete, it is an
excellent opportunity to explore the complete design and how it will address
the needs of extreme scale platforms.  This paper not only provides a timely
summary of important aspects of the design specifications but also capture the
underlying reasoning that is not available elsewhere. With this information, we
encourate the broader community to understand the design, intent, and future
directions to foster discussion guiding phase 2 and the ultimate production
storage stack based on this work.

While many of the design elements are best practices from current research,
there are many new contributions as well as different integration approaches
offering different performance and quality of service guarantees. The specifics
of how the different layers interact offer a new approach for managing data for
extreme scale platforms and workloads.

For the current demonstration system, the HDF5 user API gives a familiar
way to interact with all of the new functionality while largely insulating 
end-users from the intracisies of dealing with the advanced functionality
offered by the stack.  The various layers below the user API include optional
components that may aid in scalability and simplify implementation across a
wide range of platforms.  The data storage layer offers support both
for local machine and data center wide storage arrays as part of the design.

In addition to these features supporting HPC platforms, the stack also
addresses the need of big data applications as demonstrated through the
Arbitrary Connected Graphs tool.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform.
Adding to the complexity of the problem is the variety of Big Data problems
that are taking advantage of these platforms. Trying to address both of these
domains is a daunting problem.

To address these challenges, a joint effort between the US Department of
Energy's Office of Advanced Simulation and Computing and Advanced Scientific
Computing Research commissioned an effort to develop a design and prototype for
an IO stack suitable for the extreme scale environment called the Fast Forward
Storage and IO (FFSIO) project. This is a joint effort led by Lawrence
Livermore National Laboratory, with the DOE Data Management Nexus leads Rob
Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL.  Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. While an idealized potential system would
be the perfect target architecture, the reality of budgets has tempered many of
the decisions. For exmaple, extensive availability of NVRAN or SSDs on all of
the compute nodes is currently not economically feasible limiting some of the
potential design choices.  With this in mind, the second phase will refine this
design incorporating fault recovery and other features missing from the first
phase.

The complete design seeks to offer high availability, byte-granular,
multi-version concurrency control. Through the use of a copy-on-write style
mechanism reducing space and time requirements, multiple versions of an object
can be stored efficiently. By assuming the client interface will be through an
IO library, a more complicated interface that offers richer functionality can
be incorporated while requiring only minimal end-user code changes.  Managing
most data access in a platform-local layer rather than requiring writing to
centralized storage will better support the performance and energy requirements
of extreme scale integrated application workflows.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/arch-mapping}
\caption{Target Architecture and Component Mapping}
\label{fig:arch-mapping}
\end{figure}

At a more detailed view, the various layers of the IO stack each contribute
different functionality.  The architecture (Figure~\ref{fig:arch-mapping})
incorporates five layers, some of which have potentially optional components.
The top layer is generally a high level IO library, such as the demonstration
HDF-5 library~\cite{hdf5}. It is in dark blue. The intent is to only have
access to the storage stack through such an API to manage the complexity of
working with the lower layers and to enable advanced functionality that would
require more direct user intervention. This layer incorporates the necessary
features for ACGs from a end-user's perspective.

@@

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product.  These ideas are presented to give a
fuller picture of where the project is going rather than dwelling on any
limitations of the published documents. This is most important to illustrate
how different concepts will work across layers since that information is spread
across multiple documents and may lack a cohesive overall view.

The rest of the paper is organized as follows. A brief overview of related work
is presented first in Section~\ref{sec:related}. Section~\ref{sec:end-user}
discusses the programmatic interface end users will see when interacting with
the storage array. This will be discussed in the contect of the HDF-5 based
example library used for the functionality demonstration. Section~\ref{sec:iof}
briefly discusses the motivation and proposal for the IO forwarding layer.
Section~\ref{sec:iod} describes the IO Dispatcher layer and the broad
functionality it offers. This will detail the pieces of the layer that are
potentially optional and mention the cross-cutting features discussed in a
later, cross-cutting section. Section~\ref{sec:daos} discusses how the
persistent storage array interface itself functions. As with the IOD layer, the
disucssion of cross-cutting features will be mentioned, but disucssed more
fully in the cross-cutting section. The VOSD layer is disucssed in
Section~\ref{sec:vosd}. In particular, the mapping between the DAOS and VOSD
layers are explored as it pertains to the physical storage. Next is an
exploration of cross-cutting features like transactions and metadata management
in Section~\ref{sec:broader}. Since these and other features are spread across
multiple layers, it makes more sense to discuss them independently once an
understanding of the overall structure has been presented.  A demonstration
of the functionality is presented in Section~\ref{sec:evaluation}. This shows
that the prototype functionality validation prototype can function. The paper
is concluded in Section~\ref{sec:conclusion} with a summary of the broad issues
covered in the paper.

@@

Many projects over the last couple of decades have sought to address some
challenging aspect of parallel file system design. The recent rise of the ``Big
Data'' applications with different characteristic IO patterns have somewhat
complicated the picture. Extreme scale machines will be expected to handle both
the traditional simulation-related workloads as well as applications more
squarely in the Big Data arena. This will require some adjustments to the
underlying system for good performance for both scenarios.

@@

Since the proposal specifies a high-level IO API will be the primary end-user
interface for programmatically interacting with the FFSIO stack, the team used
the HDF-5 API and leveraged the Virtual Object Layer (VOL) for the initial
design and implementation demonstration. This also serves as a good test
determining what are strictly necessary extensions to an existing IO API to
support the new functionality.  The additional functionality, such as
transactions, can be ignored for legacy implementations, but these applications
will not be able to take advantage of the asynchronous IO support inherent to
the new API.  The additions comprise (Figure~\ref{fig:vol-arch}):

@@

%\item
%  Virtual Object Layer (VOL) offers the ability to plug in new implementations
%  for HDF-5 semantics. In this case, the HDF-5 calls are implemented using the
%  IOD or DAOS API rather than MPI-IO or POSIX.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5 files}
\label{fig:hdf5-sample}
\end{figure}

@@

\item
  \textbf{File}: In the HDF5 data model the container of an HDF5 infoset
  is represented by a file. It is a collection of objects that also
  explains the relationship between them. Every file begins with a root
  group ``/'', which serves as the ``starting-point'' in the object
  hierarchy.

\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets, datatypes or attributes within
  it.

\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays with potentially multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.

\item
  \textbf{Attribute}: Attributes are used for annotating datasets,
  groups, and datatype objects. They are datasets themselves, and are
  attached to existing objects they annotate.

@@

For example, as shown in Figure \ref{fig:hdf5-sample}, the file Sample.h5
contains the root group which itself contains a group G1 and two datasets, D1
and D2. Group G1 contains a dataset D3. Attribute A1 is linked to dataset D1.
The objects and the relationships between them can be represented as a B-tree,
which is used internally by HDF5 to index its objects.

@@

The Virtual Object Layer is an abstraction mechanism internal to the HDF5
library~\cite{hdf5}. As shown in Figure~\ref{fig:vol-arch} it is implemented
just below the public API. The VOL exports an interface that allows writing
plugins for HDF5, thereby enabling developers to handle the data in ways other
than necessarily writing to storage in an HDF5 format.  Plugin writers provide
an implementation for a set of functions that are trusted to provide the
proper semantics represented by the function for the new environment. For
example, data staging could be implemented in the VOL layer by replacing
writing to disk in the HDF5 format to sending data to a data staging area using
some messaging mechanism.

For this project, rather than the default writing to disk in the HDF-5 format,
the VOL layer is used to interact with the IOD layer and the different concepts
it offers without requiring all of the functionality be exposed to users. For
example, the containers and objects concept is transparently mapped to the
files and datasets existing HDF5 users are familiar with.  This reduces the
difficulty in porting applications to use the new IO stack.

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism}
\label{fig:vol-arch}
\end{figure}

@@

The IOD VOL plugin serves as the bridge between HDF5 and the IOD or DAOS Layer
(Figure~\ref{fig:vol-arch}). The application calls the HDF5 library while
running on the system's compute nodes. Using the VOL architecture, the IOD VOL
plugin uses a function shipper (RPC library) to forward the VOL calls to a
server component running on the IO nodes (IONs). This function shipping is the
IO Forwarding Layer discussed briefly in Section~\ref{sec:iof}. Once the calls
arrive at the IO nodes, they are translated into I/O Dispatcher (IOD) API calls
and executed at the IONs.

@@

\begin{table}[ht]
    %\vspace{-0.15in}
    \centering
    \caption[HDF5 to FFSIO Mapping]{HDF5 Data Model to FFSIO Data Model Mapping}
    \bigskip
    %\vspace{-0.15in}

@@

The IO Forwarding layer offers a mechanism to reduce the concurrency impact of
the massive process count on the storage stack. A current trend mixing MPI with
a threading library like OpenMP is addressing the same issue. In this case,
the storage stack has far fewer end-points in which to receive requests and
data from compute processes. By reducing the number of simultaneous requests,
delays can be reduced. This has been demonstrated for the file open operation
with Lustre~\cite{lofstead:2009:adaptable} and to some degree for accessing the
storage devices themselves~\cite{lofstead:2010:io-variability}. The BlueGene
platform incorporated dedicated hardware to perform this role. The proposed
functionality for this layer, beyond managing the number of connections to
the IOD layer, is to implement function shipping from the compute nodes to the
IO nodes.

@@

Strictly speaking, the IO Dispatcher layer and included functionality, such as
burst buffers, is an optional component of the proposed IO stack. All of the
functionality, such as function and analysis shipping, transaction management,
and managing asynchronous data movement can be handled by other portions of the
stack. For an extreme scale platform, the IOD layer will be an essential
pressure relief valve for the underlying persistent storage layer. By making it
optional, the proposed stack can be deployed more easily on smaller clusters or
for those with more constrainted budgets. For simplicity, the rest of this
section will describe a full stack including all of the proposed IOD
components.

The core idea for IOD is to provide a way to manage the IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
offloading the communication load from the compute nodes. IOD has three main
purposes. First, the burst buffers works as a fast cache absorbing write
operations for the slower trickle out to the central storage array. It can also
be used to retrieve objects from the central storage array for more efficient
read operations and offers data filtering to make client reads more efficient.
Second, it offers the transaction mechanism for controlling data set visibility
and to manage faults that would prevent a data set from being used. These
transactions are local to the IOD layer until persisted to the DAOS layer
eliminating the need for burdening the persistent storage with transient data.
Third, data processing operations can be placed in the IOD. These operations
are intended to offer data rearrangement, filtering, and similar operations
prior to data reaching the central storage array.

@@

%\subsection{I/O Dispatcher API}
%\label{sec:fast-forwards-io-dispatcher-api}
%
%The Fast Forward I/O and Storage project is aimed at integrating many of
%the services existing currently in middleware libraries in a
%next-generation storage stack, exposed through a non-POSIX,
%transactional, asynchronous, object-based API.
%
%From a high-level point of view, the Fast Forward's I/O Dispatcher (IOD)
%interface allows the user to:
%
%\begin{itemize}
%\itemsep1pt\parskip0pt\parsep0pt
%\item
%  transactions
%\item
%  asynchrony
%\item
%  objects
%\item
%  layout
%\item
%  placement
%\item
%  formats
%\end{itemize}
%
%Type of objects:

@@

For both IO performance and to aid in analysis and other data processing, the
multi-dimensional array object can be split across multiple IO nodes. Each
piece of this array is called a {\em shard}. The IOD API supports the following
sharding strategies:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \textbf{contiguous}. fixed chunking, distributed in a round-robin
  fashion accross the I/O nodes.
\item
  \textbf{chunked}. same as above but with irregular (sparse) chunking.
\item
  \textbf{user-defined}. either contiguous or chunked but user specifies
  where to place each individual shard.
\end{itemize}

It is possible to request the transformation of an object's physical
layout to other formats, having multiple copies of the same objects in
multiple formats if desired. Also, the user can pre-fetch objects from
the storage cluster into the I/O nodes or read them directly from the
storage cluster. At the semantic level (HDF5), indices can be created
for datasets resulting in being able to read through an index instead of
directly from the base array.

@@

\subsubsection{Sharding}
\label{sec:sharding}

The idea of sharding is to store a logically complete portion of a data set on
a single storage target. This is similiar in concept to the HDF-5 hyperslab.
The FFSIO stack supports sharding the data in the default or some other
structured way as well as ``re-sharding'' based on application needs. For
example, reordering the data so that a different dimension is the ``fast''
dimension may greatly improve the performance of a subsequent data analytics
task. A common scenario where this is common is a Fortran code (column-major)
writes data for a C code (row-major) to analyze.

Additionally, sharding at the IOD layer is done in a couple of different ways.
When writing from the compute nodes, the default layout will reflect what the
processes write. This can be reorganized using function shipping. At the DAOS
layer, this data organization may change based on the type of storage devices
beneath it.

\subsection{I/O Nodes}
IOD processes are hosted on the IO nodes that interface a general compute area
with the storage array. The IO noes handle requests forwarded by the scientific
applications, potentially integrate a tier of solid-state devices to absorb the
burst of random operations, and organize/re-format the data so that transfers
to/from the staging area from/to the traditional parallel file system can be
done more efficiently. It also has the capacity to executing analysis on the
fresh data that has just been generated by simulation applications running at
the compute nodes. As the data arrives, re-organization and data preparation
can be applied in order to anticipate the execution of analytical tasks.

\begin{figure}[htbp]
\centering
\includegraphics[width=\columnwidth]{images/exa-arch.png}
\caption{Exascale Architecture}
\label{fig:exa-arch}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure~\ref{fig:exa-arch}. The designated IO nodes (IONs) are connected to the
compute nodes (CNs) through the same fast fabric (e.g., InfiniBand), while the
connection to the external storage cluster is through a secondary slower
channel (e.g., 10Gb Ethernet). By providing additional storage, such as SSDs on
the IO nodes, these nodes are capable of better regulating the IO pressure on
the underlying storage array better than simple forwaring gateways. For this
project, using something like SSDs on the IO nodes is termed a {\em Burst
Buffer} and is disucssed below.

%In current proposals, the stack is managed by middleware that sits between the
%application and the parallel file system. Since most of the applications assume
%a POSIX interface to storage, existing middleware either transparently handle
%the I/O operations, appearing as regular POSIX calls to applications; or modify
%the I/O API as little as possible in order to minimize the impact on production
%codes.
%
%As we move towards the exascale goal, this way of interfacing with storage
%becomes more of an obstacle. Many features provided by the stack are hidden by
%the middleware/POSIX layers. If these were visible to the applications, and if
%the applications were able to control how the underlying layers behave, they
%would benefit greatly since the domain knowledge that is available at the
%application level could be used to execute storage operations efficiently. One
%more implication of the 25-year-old POSIX API is in cumbersome middleware,
%since metadata has to be part of the same uni-dimensional byte array being
%stored.
%
%The Fast Forward Storage I/O project is aimed at merging the features of
%existing middleware into a next-generation parallel file system stack. This new
%storage API termed the I/O dispatcher (IOD) is object-based, transactional,
%asynchronous and I/O-staging-aware (Figure~\ref{fig:exa-stack} (right)).

@@

Future work will generalize this model to potentially support dynamic IO node
allocation and potentially examine the possibility of oversubscription. It will
be strictly necessary to consider shared IO nodes for cases where the number
of deployed applications exceeds the number of IO nodes. This is also under
consideration for future work. This first pass is focused on extreme scale
application runs that use the vast majority of a platform rather than a
capacity cluster where end-to-end performance is a lesser concern.

>>>>

\documentclass[conference]{IEEEtran}
\pdfpagewidth=8.5in
\pdfpageheight=11in
%\usepackage{subfig}
\usepackage{subfigure}
\usepackage[pdftex]{}
\usepackage{graphicx}
%\usepackage{todonotes}
\usepackage{listings}
\lstset {% general command to set parameter(s)
         language=C,
     basicstyle=\footnotesize,               % print whole listing small
%     keywordstyle=\color{black}\bfseries, % underlined bold black keywords
%     identifierstyle =\color{black},  % nothing happens
%     commentstyle=\color{black}\emph, % white comments
     %stringstyle=\ttfamily,          % typewriter type for strings
%     stringstyle=\color{black},       % typewriter type for strings
         tabsize=4,
         showtabs=false,
     showstringspaces=false}
%can't figure this one out for particles bandwidth
%\usepackage[caption,label]{subfig}

@@

\addtolength{\parskip}{-0.08in}

@@

\begin{abstract}
The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
project is going to have significant impact on storage systems design within
and beyond the HPC community. With phase 1 of the project complete, it is an
excellent opportunity to explore the complete design and how it will address
the needs of extreme scale platforms.  This paper examimes the overall design.
Each layer of the proposed stack is discussed in some detail along with
cross-cutting topics that impact multiple layers, such as transactions and
metadata management. We also describe the proposed features to better support
some classes of big data applications.

This paper not only provides a timely summary of important aspects of the
design specifications but also captures the underlying reasoning that is not
available elsewhere. With this information, we encourate the broader community
to understand the design, intent, and future directions to foster discussion
guiding phase 2 and the ultimate production storage stack based on this work.
An initial performance evaluation of the early prototype implementation is also
provided to validate the presented design.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve extreme scale science platform
requirements.  Adding to the problem complexity is the variety of Big Data
problems users want to address using these platforms. Trying to address both of
these domains is a daunting problem.

To address these challenges, a joint effort between the US Department of
Energy's Office of Advanced Simulation and Computing and Advanced Scientific
Computing Research commissioned a project to develop a design and prototype for
an IO stack suitable for the extreme scale environment. It will be referred to
as the Fast Forward Storage and IO (FFSIO) project. This is a joint effort led
by Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL.  Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. While an idealized potential system would
be the perfect target architecture, the reality of budgets has tempered many of
the decisions. For example, extensive availability of NVRAN or SSDs on all of
the compute nodes is currently not economically feasible limiting some of the
potential design choices.  With this in mind, the second phase will refine this
design incorporating fault recovery and other features missing from the first
phase.

The complete design seeks to offer high availability, byte-granular,
multi-version concurrency control. Through the use of a copy-on-write style
mechanism reducing storage space and writing time requirements, multiple
versions of an object can be stored efficiently. By assuming the client
interface will be through an IO library, a more complicated interface offering
richer functionality can be incorporated while requiring only minimal end-user
code changes.  Managing most data access in a platform-local layer rather than
requiring writing to centralized storage will better support the performance
and energy requirements of extreme scale application compositions.

@@

\begin{figure}[htbp]
\vspace{-0.10in}
\centering
\includegraphics[width=\columnwidth]{images/arch-mapping}
\vspace{-0.15in}
\caption{Target Architecture and Component Mapping}
\label{fig:arch-mapping}
\vspace{-0.15in}
\end{figure}

At a more detailed view, the various layers of the IO stack each contribute
different functionality.  The architecture (Figure~\ref{fig:arch-mapping})
incorporates five layers, some of which have potentially optional components.
The top layer is generally a high level IO library, such as the demonstration
HDF5 library~\cite{hdf5}. It is in dark blue. The intent is to only have
access to the storage stack through such an API to manage the complexity of
working with the lower layers and to enable advanced functionality that would
require more direct user intervention. This layer incorporates the necessary
features for ACGs from a end-user's perspective.

@@

Along with analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product.  These ideas are presented to give a
fuller picture of where the project is going rather than dwelling on any
limitations of the published documents. This is most important to illustrate
how different concepts will work across layers since that information is spread
across multiple documents and may lack a cohesive overall view.

The rest of the paper is organized as follows. A brief overview of related work
is presented first in Section~\ref{sec:related}. Section~\ref{sec:end-user}
discusses the programmatic interface end users will see when interacting with
the storage array. This will be discussed in the contect of the HDF5 based
example library used for the functionality demonstration. Section~\ref{sec:iof}
briefly discusses the motivation and proposal for the IO forwarding layer.
Section~\ref{sec:iod} describes the IO Dispatcher layer and the broad
functionality it offers. This will detail the pieces of the layer that are
potentially optional and mention the cross-cutting features discussed in a
later, cross-cutting section. Section~\ref{sec:daos} discusses how the DAOS
layer functions. As with the IOD layer, the cross-cutting features will be
mentioned, but disucssed more fully in the cross-cutting section. The VOSD
layer is disucssed in Section~\ref{sec:vosd}. In particular, the mapping
between the DAOS and VOSD layers are explored as it pertains to the physical
storage. Next is an exploration of cross-cutting features like transactions and
metadata management in Section~\ref{sec:broader}. Since these and other
features are spread across multiple layers, it makes more sense to discuss them
independently once an understanding of the overall structure has been
presented.  A demonstration of the functionality is presented in
Section~\ref{sec:evaluation}. This shows that the prototype system based on the
proposed design can function. The paper is concluded in
Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
paper.

@@

Many projects over the last couple of decades have sought to address some
challenging aspect of parallel file system design. The recent rise of ``Big
Data'' applications with different characteristic IO patterns have somewhat
complicated the picture. Extreme scale machines will be expected to handle both
the traditional simulation-related workloads as well as applications more
squarely in the Big Data arena. This will require some adjustments to the
underlying system for good performance for both scenarios.

@@

Since the proposal specifies a high-level IO API will be the primary end-user
interface for programmatically interacting with the FFSIO stack, the team used
the HDF5 API and leveraged the Virtual Object Layer (VOL) for the initial
design and implementation demonstration. This also serves as a good test
determining what are strictly necessary extensions to an existing IO API to
support the new functionality.  The additional functionality, such as
transactions, can be ignored for legacy implementations, but these applications
will not be able to take advantage of the asynchronous IO support inherent to
the new API.  The additions comprise (Figure~\ref{fig:vol-arch}):

@@

%\item
%  Virtual Object Layer (VOL) offers the ability to plug in new implementations
%  for HDF5 semantics. In this case, the HDF5 calls are implemented using the
%  IOD or DAOS API rather than MPI-IO or POSIX.

@@

%\begin{figure}[htbp]
%\vspace{-0.10in}
%\centering
%\includegraphics[width=\columnwidth]{images/hdf5.png}
%\vspace{-0.20in}
%\caption{Sample file illustrating the hierarchical structure of HDF5 files}
%\label{fig:hdf5-sample}
%\vspace{-0.25in}
%\end{figure}

@@

\item
  \textbf{File}: In the HDF5 data model, the collection of data items stored
  together is represented by a file. It is an object collection that also
  describes the relationship between them. Every file begins with a root
  group ``/'' serving as the ``starting-point'' in the object hierarchy.

\item
  \textbf{Group}: A group is an object allowing association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets, datatypes or attributes within
  it.

\item
  \textbf{Dataset}: HDF5 datasets are objects representing actual data
  or content. Datasets are arrays with potentially multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions) and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.

\item
  \textbf{Attribute}: Attributes are used for annotating HDF5 objects. They are
  datasets themselves and are attached to existing objects.

@@

%For example, as shown in Figure \ref{fig:hdf5-sample}, the file Sample.h5
%contains the root group which itself contains a group G1 and two datasets, D1
%and D2. Group G1 contains a dataset D3. Attribute A1 is linked to dataset D1.
%The objects and the relationships between them can be represented as a B-tree,
%which is used internally by HDF5 to index its objects.

@@

The Virtual Object Layer is an abstraction mechanism internal to the HDF5
library~\cite{hdf5}. As shown in Figure~\ref{fig:vol-arch} it is implemented
just below the public API. The VOL exports an interface that allows writing
plugins for HDF5 enabling developers to handle data in ways other than writing
to storage in an HDF5 format.  Plugin writers provide an implementation for a
set of functions and are trusted to provide the proper semantics for the new
environment. For example, data staging could be implemented in the VOL layer by
replacing writing to disk in the HDF5 format to sending data to a data staging
area using some messaging mechanism.

For this project, rather than the default writing to disk in the HDF5 format,
the VOL is used to interact with the IOD layer and the different concepts it
offers without requiring all of the functionality be exposed to users. For
example, the containers and objects concept is transparently mapped to the
files and datasets existing HDF5 users are familiar with.  This reduces the
difficulty porting applications to the new IO stack.

\begin{figure}[htbp]
\vspace{-0.10in}
\centering
\includegraphics[width=\columnwidth]{images/vol-arch.png}
\vspace{-0.10in}
\caption{Architectural view of the VOL abstraction mechanism}
\label{fig:vol-arch}
\vspace{-0.10in}
\end{figure}

@@

The IOD VOL plugin serves as the bridge between HDF5 and the IOD or DAOS Layer
(Figure~\ref{fig:vol-arch}). The application calls the HDF5 library while
running on the system's compute nodes. Using the VOL architecture, the IOD VOL
plugin uses a function shipper (RPC library) to forward the VOL calls to a
server component running on the IO nodes (IONs). This function shipping is the
IO Forwarding Layer discussed briefly in Section~\ref{sec:iof}. Once the calls
arrive at the IO nodes, they are translated into IO Dispatcher (IOD) API calls
and executed at the IONs.

@@

\begin{table}[ht]
    \vspace{-0.10in}
    \centering
    \caption[HDF5 to FFSIO Mapping]{HDF5 Data Model to FFSIO Data Model Mapping}
    \bigskip
    \vspace{-0.15in}

@@

The IO Forwarding layer offers a mechanism to reduce the concurrency impact of
the massive process count on the storage stack. A current trend mixing MPI with
a threading library like OpenMP is addressing the same issue, but limited to
handling the parallelism on a single node rather than multiple nodes. Projected
extreme scale platforms will have far fewer storage stack end-points per
compute process in which to receive requests and data. By reducing the number
of simultaneous requests, delays can be reduced. This has been demonstrated for
the file open operation with Lustre~\cite{lofstead:2009:adaptable} and to some
degree for accessing the storage devices
themselves~\cite{lofstead:2010:io-variability}.  The BlueGene platform
incorporated dedicated hardware to perform this role. The proposed
functionality for this layer, beyond managing the number of connections to the
IOD layer, is to implement function shipping from the compute nodes to the IO
nodes.

@@

Strictly speaking, the IO Dispatcher layer and included functionality, such as
burst buffers, is optional. All of the functionality, such as function and
analysis shipping, transaction management, and managing asynchronous data
movement can be handled by other portions of the stack. For an extreme scale
platform, the IOD layer will be an essential pressure relief valve for the
underlying persistent storage layer. By making it optional, the proposed stack
can be deployed more easily on smaller clusters or for those with more
constrainted budgets. For simplicity, the rest of this section will describe a
full stack including all of the proposed IOD components.

The core idea for IOD is to provide a way to manage IO load that is
separate from the compute nodes and the storage array. Communication intensive
activities, such as data rearrangement, can be moved to the IOD layer
reducing the number of participants and message count. IOD has three main
purposes. First, the burst buffers work as a fast cache absorbing write
operations that then trickles out to the central storage array. It can also
be used to retrieve objects from the central storage array for more efficient
read operations and offers data filtering to make client reads more efficient.
Second, it offers the transaction mechanism for controlling data set visibility
and to manage faults that could expose an incomplete or corrupt data set to
users. These transactions are local to the IOD layer until persisted to the
DAOS layer eliminating the need for burdening the persistent storage with
transient data.  Third, data processing operations can be placed in the IOD.
These operations are intended to offer functionality like data rearrangement
and filtering prior to data reaching the central storage array.

@@

%\subsection{IO Dispatcher API}
%\label{sec:fast-forwards-io-dispatcher-api}
%
%The Fast Forward IO and Storage project is aimed at integrating many of
%the services existing currently in middleware libraries in a
%next-generation storage stack, exposed through a non-POSIX,
%transactional, asynchronous, object-based API.
%
%From a high-level point of view, the Fast Forward's IO Dispatcher (IOD)
%interface allows the user to:
%
%\begin{itemize}
%\itemsep1pt\parskip0pt\parsep0pt
%\item
%  transactions
%\item
%  asynchrony
%\item
%  objects
%\item
%  layout
%\item
%  placement
%\item
%  formats
%\end{itemize}
%
%Type of objects:

@@

For both IO performance and to aid in analysis and other data processing, the
multi-dimensional array object can be split across multiple IO nodes. Each
piece of this array is called a {\em shard}.  The idea of sharding is to store
a logically complete portion of a data set on a single storage target. This is
similiar in concept to the HDF5 hyperslab.  The FFSIO stack supports sharding
the data in the default or some other structured way as well as ``re-sharding''
based on application needs. For example, reordering the data so that a
different dimension is the ``fast'' dimension may greatly improve the
performance of a subsequent data analytics task. A common scenario where this
is common is a Fortran code (column-major) writes data for a C code (row-major)
to analyze.  The IOD API supports the following sharding strategies:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \textbf{contiguous}. fixed chunking, distributed in a round-robin
  fashion accross the IO nodes.
\item
  \textbf{chunked}. same as above but with irregular (sparse) chunking.
\item
  \textbf{user-defined}. either contiguous or chunked, but user specifies
  where to place each individual shard.
\end{itemize}

It is possible to request the transformation of an object's physical
layout to other formats, having multiple copies of the same objects in
multiple formats if desired. Also, the user can pre-fetch objects from
the storage cluster into the IO nodes or read them directly from the
storage cluster. At the semantic level (HDF5), indices can be created
for datasets resulting in being able to read through an index instead of
directly from the base array.

@@

%At the DAOS layer, the sharding can be changed based on the type of storage
%devices beneath it.

\subsection{IO Nodes}
IOD processes are hosted on the IO nodes that interface a general compute area
with the storage array. The IO nodes handle requests forwarded by the
scientific applications, potentially integrate a tier of solid-state devices to
absorb the burst of random or high volume operations, and organize/re-format
the data so that transfers to/from the staging area from/to the traditional
parallel file system can be done more efficiently. It also has the capacity to
execute analysis on data recently generated by simulation applications running
at the compute nodes, but not persisted to the storage array. As the data
arrives, re-organization and data preparation can be applied in order to
anticipate the execution of analytical tasks.

\begin{figure}[htbp]
\centering
\vspace{-0.10in}
\includegraphics[width=\columnwidth]{images/exa-arch.png}
\vspace{-0.20in}
\caption{Exascale Architecture}
\label{fig:exa-arch}
\vspace{-0.10in}
\end{figure}

A common configuration for this type of deployment is shown in
Figure~\ref{fig:exa-arch}. The designated IO nodes (IONs) are connected to the
compute nodes (CNs) through the same fast fabric (e.g., InfiniBand) while the
connection to the external storage cluster is through a secondary, slower
channel (e.g., 10Gb Ethernet). By providing additional storage on the IO nodes,
such as SSDs, these nodes are capable of better regulating the IO pressure on
the underlying storage array better than simple forwaring gateways. For this
project, using something like SSDs on the IO nodes is termed a {\em Burst
Buffer} and is disucssed below.

%In current proposals, the stack is managed by middleware that sits between the
%application and the parallel file system. Since most of the applications assume
%a POSIX interface to storage, existing middleware either transparently handle
%the IO operations, appearing as regular POSIX calls to applications; or modify
%the IO API as little as possible in order to minimize the impact on production
%codes.
%
%As we move towards the exascale goal, this way of interfacing with storage
%becomes more of an obstacle. Many features provided by the stack are hidden by
%the middleware/POSIX layers. If these were visible to the applications, and if
%the applications were able to control how the underlying layers behave, they
%would benefit greatly since the domain knowledge that is available at the
%application level could be used to execute storage operations efficiently. One
%more implication of the 25-year-old POSIX API is in cumbersome middleware,
%since metadata has to be part of the same uni-dimensional byte array being
%stored.
%
%The Fast Forward Storage IO project is aimed at merging the features of
%existing middleware into a next-generation parallel file system stack. This new
%storage API termed the IO dispatcher (IOD) is object-based, transactional,
%asynchronous and IO-staging-aware (Figure~\ref{fig:exa-stack} (right)).

@@

Future work will generalize this model to potentially support dynamic IO node
allocation and examine the possibility of oversubscription. It will be strictly
necessary to consider shared IO nodes for cases where the number of deployed
applications exceeds the number of IO nodes.  This first phase focuses on
extreme scale application runs that use the vast majority of a platform rather
than a capacity cluster where end-to-end performance is a lesser concern.
